Enabling Ambient Backscatter Using Low-cost Software-defined-radio

### 2017-04-06 paper.tex
Moved bibliography to bibtex

\subsection{RTL2832U}
\label{sub:rtl2832}
\begin{figure}[h]
\centering
\includegraphics[width=0.5\columnwidth]{./fig/rtlsdr.jpg}
\caption{\textit{RTL-SDR} hardware with the DVB-T I/Q demodulator \textit{Raeltek RTL2832U} (left IC) and the tuner with integrated LNA \textit{Rafael Micro R820T/2} (right IC).}
\label{fig:receiver_arch}
\end{figure}
The Raeltek RTL2832U is a terrestrial digital video broadcast (DVB-T) chip, which is the basis of a very low-cost software defined radio: The RTL-SDR. One feature of this chip is, that it supports USB 2.0. Ham radio enthusiasts combined their efforts to enable accessing the raw I/Q data. With their software driver (basically the \textit{librtlsdr}, cf. \cite{github:librtlsdr}) DVB-T sticks, which are based on the RTL2832, can be converted into a wideband SDR. This sort of hardware has been in the range of a few thousands USD a few years ago. An RTL2832U based DVB-T receiver can be bought with antenna for less than 10 USD. 
What is a software defined radio? Traditionally basic radio components like mixers or filters have been realized in hardware. Nowadays as soon as one works in the baseband frequency the coputational resources of a personal computer are sufficient to realize these components with signal processing done in software.
An RTL-SDR does not only consist of the RTL2832U. The second essential component is the tuner, which mixes down the radio frequency (RF) to a intermediate frequency (IF). In the case of the used hardware this tuner component was represented by a \textit{Rafael Micro R820T/2}. It subsumes a low-noise amplifier (LNA) and filters. The image frequency is rejected automatically with 65 dBc and the noise figure is with 3.5 dB quite low. It alows a frequency range from 42 to 1002 MHz (cf. \cite{rafael:r820t}). 
The I/Q demodulator, also shown in figure \ref{fig:receiver_arch}, is the most important part of the \textit{RTL2832U} for our work. The received signal can be intepreted as
\begin{equation}
	s_{IF}(t)=I(t) \cdot cos(\omega_{0}t) + Q(t) \cdot sin(\omega_{0}t)
\end{equation}     
This is multiplied with a cosine (and a sine as well) from a local oscillator.
\begin{multline}
        s_{IF}(t) \cdot s_L(t) = I(t) \cdot cos(\omega_{0}t) \cdot cos(\omega_{0}t)\\+ Q(t) \cdot sin(\omega_{0}t) \cdot cos(\omega_{0}t)
\end{multline}
With \ensuremath{2\,cos(a)cos(b)=cos(a-b)+cos(a+b)} and \ensuremath{2\,sin(a)cos(b)=sin(a+b)+sin(a-b)} follows
\begin{multline}
        2 \cdot s_{IF}(t) \cdot s_L(t) = I(t) \cdot \bigl[1+cos(2\omega_{0}t)\bigr]\\+ Q(t) \cdot \bigl[sin(2\omega_0t)+sin(0)\bigr].
\end{multline}
One can see, that the interesting inphase part in this case is represented by a DC value after mixing. With a low-pass filter this DC value can be seperated from the undesired rest. Similar things are hapening with the quadrature part when mixing with a sine, which is indeed done to get the quadrature part.  
\section{Design}

\subsection{Transmitter}
\begin{figure}[h]
\centering
\includegraphics[width=0.8\columnwidth]{./fig/s11v2}
\caption{Input reflection coefficient \ensuremath{\Gamma_{in}} also known as \ensuremath{S_{\text{11}}} of the selfmade backsacttering ground plane antenna in dB.}
\label{fig:s11}
\end{figure}
A backscattering antenna is the most important part of backscattering sender. Our antenna is optimized for usage with the television signal, which has its center frequency at 626 MHz. It is a ground plane antenna, which has been built up on a big PCB. This rectangular PCB, with dimensions of 200 mm x 300 mm, is covered with a copper layer on one side and left blank on the other. A hole in the midle of the PCB serves for connecting a thicker wire to a 50 \ensuremath{\Omega} microstrip transmission line. The wire is the vertical antenna element has a length of 350 mm, which is roughly 0.73 \ensuremath{\lambda}. It is soldered to the transmission line. The transmission line has been calculated for 626 MHz using FR4 substrate with a thickness of 1.6 mm. According to microstrip line theory the length of the line changes the phase and the width of the line changes the impedance (cf. chapter 3 in \cite{book:pozar}). The design equations given in \cite{books:pozar} are implemented in various programs e.g. KiCAD, which has been used for solving the design equations. The given parameters lead to a microstrip line width of 3 mm. An open-stub, attached directly at the puncture where the vertical antenna element is attached to the transmission line, is used for fine tuning of the input impedance. The stub length has been adjusted manually (shortened) while measuring the reflection coefficient with a vector network analyzer (VNA). Open-stub matching is discussed in chapter 5 in \cite{book:pozar}. Furthermore the resonance frequency has been tuned by cutting the vertical antenna element piece by piece while measuring with the VNA. At some point the \ensuremath{S_{\text{11}}} had the minimum quite close to where we wanted to have it (cf. \ref{fig:s11}). In the plot one can also see, that the antenna is quite narrowband. For backscattering this is not necessarily a disadvantage, since everything, which is within the operating frequency of the antenna is backscattered. So with a narrowband antenna the effect, that signals are backscattered, which are not targeted, is reduced. This statement has to be qualified as the antenna also has multiple resonance frequencies in multiple bands. With our approach we were able to achieve a input resistance at 634 MHz of \ensuremath{Z_{in}=(52.2 + j 4.3) \Omega}. At 626 MHz \ensuremath{S_{\text{11}}} is still approximately -17 dB.                   
Besides the antenna the transmitter consists out of a \textit{MSP430}, which acts as subscriber, that sends data to the receiver. This transmitter is connected with one I/O pin and ground to a rf switch. The switch can change between matched (50 \ensuremath{\Omega}) and open. The software now steers this switch with a two 2 MHz rectangular timer signal for a 1 and is just leavs it open for a 0. Therefore the result is a shift of the television signal by 2 MHz for a 1 and no shift for a 0. Hence a classical binary amplitude shift keying is the implemented modulation technique.

@@

The receiver was the most work intensive part of the design. C++ has been used to implement the receiver. We provide the code publicly available under \url{https://github.com/s3xm3x/backscatterBASKReceiver}. In figure \ref{fig:receiver_arch} one can see the architecture of the receiver from an abstarct point of view. A highly sophisticated bus system has been developed to exchange data between the different components of the system. The signal flow is from the left to the right. \textit{Librtlsdr} (cf. \cite{github:librtlsdr}), which is based on \textit{libusb}, is used to transfer the data from the TV stick into our program. As described under \ref{sub:rtl2832}, the RTL2832 mixes down the high frequency to a intermediate frequency (both values are in SW adjustable). Subsequently the data is mixed with sine and cosine, low-pass filtered and finally transfered in the digital world. This data is available as two 8 Bit values. It is a well-known fact, that this technique, which is commonly known as I/Q demodulation, results in a real and an imagenery part. With this two values phase and absolute value of the data can be determined for every sample. So every sample consists out of a real and and imaginary part. And can be seen as:

@@

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
\begin{thebibliography}{1}

\bibitem{github:librtlsdr}
Steve Markgraf et. al. \emph{librtlsdr}. \url{https://github.com/steve-m/librtlsdr}. Github: 2012-2017.

\bibitem{rafael:r820t}
\emph{High Performance Low Power Advanced Digital TV Silicon Tuner}. R820T. Rev 1.2. Rafael Micro. November 2011.

\bibitem{book:pozar}
D. M. Pozar. \emph{Microwave Engineering}. Fourth Edition. John Wiley \& Sons Inc. 111 River Street, Hoboken. 2012

\end{thebibliography}

>>>>

\subsection{RTL2832U}
\label{sub:rtl2832}
\begin{figure}[h]
\centering
\includegraphics[width=0.5\columnwidth]{./fig/rtlsdr.jpg}
\caption{\textit{RTL-SDR} hardware with the DVB-T I/Q demodulator \textit{Raeltek RTL2832U} (left IC) and the tuner with integrated LNA \textit{Rafael Micro R820T/2} (right IC).}
\label{fig:receiver_arch}
\end{figure}
The Raeltek RTL2832U is a terrestrial digital video broadcast (DVB-T) chip, which is the basis of a very low-cost software defined radio: The RTL-SDR. One feature of this chip is, that it supports USB 2.0. Ham radio enthusiasts combined their efforts to enable accessing the raw I/Q data. With their software driver (basically the \textit{librtlsdr}, cf. \cite{steve-m_librtlsdr}) DVB-T sticks, which are based on the RTL2832, can be converted into a wideband SDR. This sort of hardware has been in the range of a few thousands USD a few years ago. An RTL2832U based DVB-T receiver can be bought with antenna for less than 10 USD. 
What is a software defined radio? Traditionally basic radio components like mixers or filters have been realized in hardware. Nowadays as soon as one works in the baseband frequency the coputational resources of a personal computer are sufficient to realize these components with signal processing done in software.
An RTL-SDR does not only consist of the RTL2832U. The second essential component is the tuner, which mixes down the radio frequency (RF) to a intermediate frequency (IF). In the case of the used hardware this tuner component was represented by a \textit{Rafael Micro R820T/2}. It subsumes a low-noise amplifier (LNA) and filters. The image frequency is rejected automatically with 65 dBc and the noise figure is with 3.5 dB quite low. It alows a frequency range from 42 to 1002 MHz (cf. \cite{rafael_r820t_2011}). 
The I/Q demodulator, also shown in figure \ref{fig:receiver_arch}, is the most important part of the \textit{RTL2832U} for our work. The received signal can be intepreted as
\begin{equation}
	s_{IF}(t)=I(t) \cdot cos(\omega_{0}t) + Q(t) \cdot sin(\omega_{0}t)
\end{equation}     
This is multiplied with a cosine (and a sine as well) from a local oscillator.
\begin{multline}
        s_{IF}(t) \cdot s_L(t) = I(t) \cdot cos(\omega_{0}t) \cdot cos(\omega_{0}t)\\+ Q(t) \cdot sin(\omega_{0}t) \cdot cos(\omega_{0}t)
\end{multline}
With \ensuremath{2\,cos(a)cos(b)=cos(a-b)+cos(a+b)} and \ensuremath{2\,sin(a)cos(b)=sin(a+b)+sin(a-b)} follows
\begin{multline}
        2 \cdot s_{IF}(t) \cdot s_L(t) = I(t) \cdot \bigl[1+cos(2\omega_{0}t)\bigr]\\+ Q(t) \cdot \bigl[sin(2\omega_0t)+sin(0)\bigr].
\end{multline}
One can see, that the interesting inphase part in this case is represented by a DC value after mixing. With a low-pass filter this DC value can be seperated from the undesired rest. Similar things are hapening with the quadrature part when mixing with a sine, which is indeed done to get the quadrature part.  
\section{Design}

\subsection{Transmitter}
\begin{figure}[h]
\centering
\includegraphics[width=0.8\columnwidth]{./fig/s11v2}
\caption{Input reflection coefficient \ensuremath{\Gamma_{in}} also known as \ensuremath{S_{\text{11}}} of the selfmade backsacttering ground plane antenna in dB.}
\label{fig:s11}
\end{figure}
A backscattering antenna is the most important part of backscattering sender. Our antenna is optimized for usage with the television signal, which has its center frequency at 626 MHz. It is a ground plane antenna, which has been built up on a big PCB. This rectangular PCB, with dimensions of 200 mm x 300 mm, is covered with a copper layer on one side and left blank on the other. A hole in the midle of the PCB serves for connecting a thicker wire to a 50 \ensuremath{\Omega} microstrip transmission line. The wire is the vertical antenna element has a length of 350 mm, which is roughly 0.73 \ensuremath{\lambda}. It is soldered to the transmission line. The transmission line has been calculated for 626 MHz using FR4 substrate with a thickness of 1.6 mm. According to microstrip line theory the length of the line changes the phase and the width of the line changes the impedance (cf. chapter 3 in \cite{pozar_microwave_2011}). The design equations given in \cite{pozar_microwave_2011} are implemented in various programs e.g. KiCAD, which has been used for solving the design equations. The given parameters lead to a microstrip line width of 3 mm. An open-stub, attached directly at the puncture where the vertical antenna element is attached to the transmission line, is used for fine tuning of the input impedance. The stub length has been adjusted manually (shortened) while measuring the reflection coefficient with a vector network analyzer (VNA). Open-stub matching is discussed in chapter 5 in \cite{pozar_microwave_2011}. Furthermore the resonance frequency has been tuned by cutting the vertical antenna element piece by piece while measuring with the VNA. At some point the \ensuremath{S_{\text{11}}} had the minimum quite close to where we wanted to have it (cf. \ref{fig:s11}). In the plot one can also see, that the antenna is quite narrowband. For backscattering this is not necessarily a disadvantage, since everything, which is within the operating frequency of the antenna is backscattered. So with a narrowband antenna the effect, that signals are backscattered, which are not targeted, is reduced. This statement has to be qualified as the antenna also has multiple resonance frequencies in multiple bands. With our approach we were able to achieve a input resistance at 634 MHz of \ensuremath{Z_{in}=(52.2 + j 4.3) \Omega}. At 626 MHz \ensuremath{S_{\text{11}}} is still approximately -17 dB.                   
Besides the antenna the transmitter consists out of a \textit{MSP430}, which acts as subscriber, that sends data to the receiver. This transmitter is connected with one I/O pin and ground to a rf switch. The switch can change between matched (50 \ensuremath{\Omega}) and open. The software now steers this switch with a two 2 MHz rectangular timer signal for a 1 and is just leavs it open for a 0. Therefore the result is a shift of the television signal by 2 MHz for a 1 and no shift for a 0. Hence a classical binary amplitude shift keying is the implemented modulation technique.

@@

The receiver was the most work intensive part of the design. C++ has been used to implement the receiver. We provide the code publicly available under \url{https://github.com/s3xm3x/backscatterBASKReceiver}. In figure \ref{fig:receiver_arch} one can see the architecture of the receiver from an abstarct point of view. A highly sophisticated bus system has been developed to exchange data between the different components of the system. The signal flow is from the left to the right. \textit{Librtlsdr} (cf. \cite{steve-m_librtlsdr}), which is based on \textit{libusb}, is used to transfer the data from the TV stick into our program. As described under \ref{sub:rtl2832}, the RTL2832 mixes down the high frequency to a intermediate frequency (both values are in SW adjustable). Subsequently the data is mixed with sine and cosine, low-pass filtered and finally transfered in the digital world. This data is available as two 8 Bit values. It is a well-known fact, that this technique, which is commonly known as I/Q demodulation, results in a real and an imagenery part. With this two values phase and absolute value of the data can be determined for every sample. So every sample consists out of a real and and imaginary part. And can be seen as:

@@

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
\bibliography{paper}
\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
%\begin{thebibliography}{1}
%
%\bibitem{github:librtlsdr}
%Steve Markgraf et. al. \emph{librtlsdr}. \url{https://github.com/steve-m/librtlsdr}. Github: 2012-2017.
%
%\bibitem{rafael:r820t}
%\emph{High Performance Low Power Advanced Digital TV Silicon Tuner}. R820T. Rev 1.2. Rafael Micro. November 2011.
%
%\bibitem{book:pozar}
%D. M. Pozar. \emph{Microwave Engineering}. Fourth Edition. John Wiley \& Sons Inc. 111 River Street, Hoboken. 2012
%
%\end{thebibliography}
